---
title: "Analysis 2: Predictive Modeling for House Prices"
description: "This analysis focuses on building predictive models to forecast future trends based on historical data.
author: "Aaron Powell & Joshua Samuel"
date: r.Sys.Date()
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(car)
source("code/utils_data_prep.R") # Data cleaned and prepped for LR
```

## Initial Multiple Linear Regression review

We will proceed with log(SalePrice) since Analysis 1 displayed some heteroscedasticity. 

```{r}
full_model <- lm(log(SalePrice) ~ ., data = train)
summary(full_model)
```

The summary of the full model NA values for several predictors. This is likely due to perfect multicollinearity (e.g., YearBuilt and YearRemodAdd are highly correlated). 

```{r}
full_model <- lm(log(SalePrice) ~ . - YearRemodAdd, data = train) # found YearRemodAdd perfectly collinear with YearBuilt
summary(full_model)
vif_values <- vif(full_model)
print(vif_values[vif_values > 5]) # start with the most extreme VIF values and work down
```



# - Create key derived features (examples relevant to our course):
#     - log_SalePrice (for modeling and Kaggle submission)
#     - HouseAge = YrSold - YearBuilt
#     - TotalSF = GrLivArea + TotalBsmtSF
#     - TotalBath = FullBath + HalfBath + BsmtFullBath*0.5 + BsmtHalfBath*0.5
#     - Quality interactions or dummy variables for high-impact categoricals (e.g., OverallQual, Neighborhood, ExterQual, KitchenQual)
#     - Any other feature creation justified by domain knowledge or EDA (e.g., log transformations on skewed continuous vars like LotArea, GrLivArea).
# - Filter or subset if needed, but remember: Analysis 2 uses ALL neighborhoods.

# 3. Exploratory Data Analysis (EDA) for Modeling --------------------
# - Summary statistics and correlations (focus on SalePrice and key predictors).
# - Visualizations: scatterplots (SalePrice vs. GrLivArea, colored by key factors), histograms of SalePrice (and log(SalePrice)), boxplots by categorical vars.
# - Identify potential transformations and interactions early.

# 4. Model 1: Simple Linear Regression (Baseline) ---------------------
# - Formula: log(SalePrice) ~ GrLivArea   (or SalePrice if you prefer; discuss choice)
# - Fit with lm()
# - Store model object (e.g., model_slr)
# - Basic summary and initial diagnostics

# 5. Model 2: Required Multiple Linear Regression (MLR1) --------------
# - Formula: log(SalePrice) ~ GrLivArea + FullBath
# - Fit with lm()
# - Store as model_mlr1

# 6. Model 3+: Enhanced Multiple Linear Regression Models -------------
# - Build at least one (ideally 2–3) improved models using techniques from class:
#     - Add meaningful predictors (e.g., OverallQual, HouseAge, TotalSF, Neighborhood as factor/dummies, ExterQual, etc.)
#     - Include transformations (log(GrLivArea), sqrt(LotArea), etc.)
#     - Add interactions (e.g., GrLivArea * OverallQual, GrLivArea * Neighborhood, YearBuilt * Remod)
#     - Feature selection: forward/backward/stepwise (stepAIC or manual), or correlation/VIF screening
# - Example enhanced model: log(SalePrice) ~ log(GrLivArea) + OverallQual + HouseAge + TotalBath + Neighborhood + GrLivArea:OverallQual + ...
# - Fit each as model_mlr2, model_mlr3, etc.
# - Comment heavily on why each variable/interaction was added.

# 7. Model Diagnostics and Assumption Checking (Critical!) ------------
# For EVERY model (SLR, MLR1, and enhanced):
#   a. Residual plots:
#      - Residuals vs. Fitted (check linearity & homoscedasticity)
#      - Normal Q-Q plot (normality of residuals)
#      - Scale-Location plot
#      - Residuals vs. Leverage (outliers/influential points)
#   b. Influential point analysis:
#      - Cook's Distance (> 4/n threshold)
#      - Leverage (hat values > 2p/n)
#      - DFFITS, DFBETAS if needed
#   c. Multicollinearity: VIF (variance inflation factors) – address if > 5–10
#   d. Address violations:
#      - Transformations already in model?
#      - Remove influential points? (justify with sensitivity analysis)
#      - Add polynomial terms or interactions if curvature present?
# - Create diagnostic plot functions or loops to avoid repetition.

# 8. Model Comparison and Evaluation ----------------------------------
# - Create a comparison table (as required in the project description):
#   Columns: Model, Adjusted R², CV PRESS (use PRESS statistic or cvTools/ caret for cross-validation), AIC, Kaggle Score (RMSE on log scale)
# - Compute internal CV PRESS (e.g., via leave-one-out or k-fold)
# - Generate predictions on test set for Kaggle:
#     - Predict log(SalePrice), exponentiate back to SalePrice
#     - Create submission.csv with Id and SalePrice
#     - Note the public Kaggle score for each model
# - Clearly state which model is "best" and why (balance of fit, assumptions, predictive performance, parsimony).

# 9. Best Model Selection and Kaggle Submission -----------------------
# - Export the final submission file(s)
# - Optional: Save model objects (RDS) for Shiny app or report reproducibility

# 10. Session Info and Reproducibility --------------------------------
# sessionInfo()
# Save any plots/tables for the report appendix

# ========================================================
# End of Script
# Remember: All code must be well-commented and runnable by the instructor.
# ========================================================
