---
title: "Analysis 2: Predictive Modeling for House Prices"
description: "This analysis focuses on building predictive models to forecast future trends based on historical data."
author: "Aaron Powell & Joshua Samuel"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    df-print: kable
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(GGally)
library(car)
library(purrr)
library(here)
source(here("code", "utils_data_prep.R"))
```

# Introduction

This analysis focuses on building predictive models for house prices using linear regression techniques, starting with simple linear regression and progressing to more complex multiple linear regression models.

## Simple Linear Regression

Taking a simple approach first, we will model log(SalePrice) as a function of Above grade (ground) living area square feet (GrLivArea). GrLivArea is a strong predictor of SalePrice and is commonly used in housing price models.

```{r}
# Compute correlation matrix from numeric variables only
numeric_train <- train |> select(where(is.numeric))
cor_matrix <- cor(numeric_train, use = "complete.obs")

saleprice_cor <- cor_matrix[, "SalePrice"]
cor_df <- tibble(
  Variable = names(saleprice_cor),
  Correlation = round(saleprice_cor, 3),
  Abs_Correlation = abs(Correlation)
) |>
  arrange(desc(Abs_Correlation))
print("Top variables by |correlation| with SalePrice:")
print(cor_df, n = 10)
```

```{r}
model_slr <- lm(log(SalePrice) ~ GrLivArea, data = train, subset = c(-1299, -524)) # Exclude outlier 1299 and 524
summary(model_slr)
par(mfrow = c(2, 2))
plot(model_slr)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(model_slr$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(model_slr$residuals), sd = sd(model_slr$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

1299 remains an outlier. It is a 5600 sqft house sold as a unfinished build for 160,000. This is an extreme case and likely not representative of the general market, so we will exclude it from our modeling.

524 is a 4700 sqft house sold for 184750 as a partial unfinished new build, which represents another extreme case. For SLR we are not considering any other factor than GrLivArea so we will exclude this special case.

## Initial Multiple Linear Regression

Number of Full Bathrooms (FullBath) is another highly correlated predictor of SalePrice. We will add it to the model to see if it improves fit and addresses any issues with the SLR model.

```{r}
model_mlr1 <- lm(log(SalePrice) ~ GrLivArea + FullBath, data = train, subset = c(-1299, -524, -636, -1124))
summary(model_mlr1)

par(mfrow = c(2, 2))
plot(model_mlr1)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(model_mlr1$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(model_mlr1$residuals), sd = sd(model_mlr1$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

636 is a 3395 sqft house sold for 200,000 as a abnormal sale. This is a large house sold for a very low price. Since abnormal sales can be anything from forclosure to short sale or trade, we may consider excluding this group. For this model we will only exclude the one outlier. 

Only explaining another 3% of the variance, this model is not a significant improvement over the SLR. The residual plots show similar patterns to the SLR, with some improvement in homoscedasticity but still some curvature in the residuals vs fitted plot. The histogram of residuals looks more normal. 

## Enhanced Multiple Linear Regression Models

Starting with a full model including all predictors and then we will work down to a more parsimonious model. 

```{r}
full_model <- lm(log(SalePrice) ~ ., data = train)
summary(full_model)
par(mfrow = c(2, 2))
plot(full_model)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(full_model$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(full_model$residuals), sd = sd(full_model$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

Histogram of residuals looks very condensed to the center and long tailed. Clearly not normal. Studentized residuals show some curvature in the residuals vs fitted plot, but overall the model seems to be doing a good job of capturing the variance in SalePrice with the log transformation.

The model displays overfit and normality concerns. We will work on improving the model by removing some predictors and adding interactions and transformations.

### Linearity Check

First we will look for linearity among numerical predictors.

```{r, warning=FALSE, message=FALSE}
# Get numeric predictors (exclude SalePrice)

# Exclude the two extreme outliers identified earlier
train_filtered <- train |> 
  filter(!Id %in% c(524, 1299)) # These two really skew the relationships and are not representative of the general market

numeric_predictors <- train_filtered |> 
  select(where(is.numeric), -SalePrice) |> 
  names()

# Reshape to long format for faceting
train_long <- train_filtered |> 
  pivot_longer(
    cols = all_of(numeric_predictors),
    names_to = "Predictor",
    values_to = "Value"
  )

# Create the linearity diagnostic plot
ggplot(train_long, aes(x = Value, y = log(SalePrice))) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", color = "red", linewidth = 1, se = FALSE) +
  facet_wrap(~ Predictor, scales = "free_x", ncol = 4) +
  labs(
    title = "Linearity Check: Numeric Predictors vs log(SalePrice)",
    subtitle = "Red line = loess smoother. Look for straight lines (good linearity).",
    x = NULL,
    y = "log(SalePrice)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 8)
  )
```

#### Testing numerical predictors

Reviewing the scatterplots of numeric data against log(SalePrice) with loess smoothers:

GrLiveArea shows a possible quadratic relationship, with diminishing returns at a higher value. 

X1stFlrSF and X2ndFlrSf show linear relationships. But this may already be captured by GrLivArea, which is the sum of 1st and 2nd floor area. We may consider removing these two predictors to address multicollinearity.

```{r}
reduced <- lm(log(SalePrice) ~ GrLivArea, data = train_filtered)
full <- lm(log(SalePrice) ~ GrLivArea + X1stFlrSF + X2ndFlrSF, data = train_filtered)
anova(reduced, full)
vif(full)
```

While there is evidence to suggest 1stFlrSF and 2ndFlrSF improve the model in a statistical sense, multicollinearity occurs because GrLIvArea is effectively of sum of the two variables. 

TotRmsAbvGrd show possible linearity but could be capturing the same information as GrLivArea, so we may consider removing it to address multicollinearity.

```{r}
full <- lm(log(SalePrice) ~ GrLivArea + TotRmsAbvGrd, data = train_filtered)
anova(reduced, full)
vif(full)
```

There is overwhelming evidence adding TotRmsAbvGrd to a model containing GrLivArea produces meaningful predictive power(parital f-test p-value <.0001). Therefore we will keep TotRmsAbvGrd in the model. 

LotArea shows a possible linear relationship, with diminishing returns at lower values and an outlier at the high end or possible quadratic relationship at extremes. 

```{r}
full <- lm(log(SalePrice) ~ GrLivArea + LotArea, data = train_filtered)
anova(reduced, full)
vif(full)
```

There is overwhelming evidence adding LotArea produces meaningful predictive power(parital f-test p-value <.0001). Therefore we will keep LotArea in the model.

BSmtFinSf1 and TotalBsmtSF show possible linear relationship with one outlier at the high end. We may consider grouping all basement variables and testing for lack of fit against a model with TotalBsmtSF as the only basement predictor.

```{r}
full = lm(log(SalePrice) ~ GrLivArea + TotalBsmtSF, data = train_filtered)
anova(reduced, full)
vif(full)
```

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea + TotalBsmtSF, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + TotalBsmtSF + BsmtQual + BsmtExposure, data = train_filtered)
anova(reduced, full)
vif(full)
```

There's overwhelming evidence to suggest TotalBsmtSF in a model with GrLivArea has meaningful predictive power (partial f-test p-value <.0001). Any attempts to add other bsmt variables results in high multicolinearity. 

GarageArea shows a possible linear relationship with log(SalePrice) but with some outliers at the high end. We may consider grouping all garage variables and testing for lack of fit against a model with GarageArea as the only garage predictor.

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + GarageArea, data = train_filtered)
anova(reduced, full)
vif(full)
```

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea + GarageArea, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + GarageArea + GarageType, data = train_filtered)
anova(reduced, full)
vif(full)
```

There's overwhelming evidence to suggest GarageArea in a model with GrLivArea and GarageType has meaningful predictive power (partial f-test p-value <.0001). Any attempts to add other Garage variables other than GarageType results in high multicolinearity. 

FirePlaces show possible linear relationship with log(SalePrice) but may not be enough to provide predictive power. We may consider grouping all fireplace variables and testing for lack of fit against a model with FirePlaces as the only fireplace predictor.

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + Fireplaces, data = train_filtered)
anova(reduced, full)
vif(full)
```

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea + Fireplaces, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + Fireplaces + FireplaceQu, data = train_filtered)
anova(reduced, full)
vif(full)
```

There's overwhelming evidence to suggest Fireplaces in a model with GrLivArea and FireplaceQu has meaningful predictive power (partial f-test p-value <.0001). 

While WoodDeckSF has potential linear relationship, the variance is too high while linearity is nearly flat. We may consider removing this predictor.

We saw earliier in model_mlr1 that FullBath has a linear relationship with log(SalePrice) and improved our model by adding 3% to the R². We will keep it in the model unless it shows multicollinearity with other predictors.

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea + FullBath, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + FullBath + HalfBath, data = train_filtered)
anova(reduced, full)
vif(full)
```

There's evidence to suggest Halfbath may have some predictive power (partial f-test p-value = 0.0095).

Flat relationships with log(SalePrice) are observed and will be removed from our model:
BedroomAbvGr, BsmtFullBath, BsmtHalfBath, EnclosedPorch, HalfBath, KitchenAbvGr, LowQualFinSf, MiscVal, OpenPorchSF, PoolArea, ScreenPorch, X3SsnPorch, YearBuilt.

Considering these values have potential linearity, but too strong of grouping at 0, we may consider using binary predictors: MasVnrArea, WoodDeckSf

LotFrontage has imputed values that may be skewing the relationship. We will proceed with caution. 

#### Results

```{r}
Removed = union(
  Removed, c(
    'X1stFlrSF', 'X2ndFlrSF', 'BsmtQual', 'BsmtCond', 'BsmtExposure', 
    'BsmtFinType1', 'BsmtFinSF1', 'BsmtFinType2', 'BsmtFinSF2', 'BsmtUnfSF', 
    'BsmtFullBath', 'BsmtHalfBath', 'GarageFinish', 'GarageQual', 'GarageCond', 
    'WoodDeckSF', 'BedroomAbvGr', 'BsmtFullBath', 'BsmtHalfBath', 'EnclosedPorch', 
    'KitchenAbvGr', 'LowQualFinSf', 'MiscVal', 'OpenPorchSF', 'PoolArea', 
    'ScreenPorch', 'X3SsnPorch', 'YearBuilt'
  )
)
```

#### Testing categorical predictors

```{r}
library(patchwork)

# Identify all categorical predictors
cat_vars <- train_filtered |> 
  select(where(~is.factor(.) || is.character(.))) |> 
  names()

# Function to visualize one categorical predictor
plot_categorical <- function(var_name) {
  train_filtered |> 
    ggplot(aes(
      x = reorder(.data[[var_name]], log(SalePrice), FUN = median),
      y = log(SalePrice)
    )) +
    geom_boxplot(outlier.alpha = 0.6) +
    labs(
      title = paste("log(SalePrice) by", var_name),
      subtitle = paste("Levels:", length(unique(train_filtered[[var_name]]))),
      x = NULL,
      y = "log(SalePrice)"
    ) +
    theme_bw(base_size = 12)
}

wrap_plots(map(cat_vars[1:6], plot_categorical), ncol = 2)
wrap_plots(map(cat_vars[7:12], plot_categorical), ncol = 2)
wrap_plots(map(cat_vars[13:18], plot_categorical), ncol = 2)
wrap_plots(map(cat_vars[19:24], plot_categorical), ncol = 2)
wrap_plots(map(cat_vars[25:30], plot_categorical), ncol = 2)
wrap_plots(map(cat_vars[31:36], plot_categorical), ncol = 2)
```

We'll focus on variables that may not have a strong relationship with SalePrice to consider removing from the model. We've confirmed GrLivArea, GarageArea, TotalBsmtSF, and FullBath have strong relationships with SalePrice so we will test other categorical predictors against a reduced model with these four predictors.

```{r}
# Ordered evaluation of all categorical predictors by RSS drop
evaluate_all_categorical <- function() {
  base_formula <- "log(SalePrice) ~ GrLivArea + GarageArea + TotalBsmtSF + FullBath"
  
  results <- map_dfr(cat_vars, function(var_name) {
    full_formula <- as.formula(paste(base_formula, "+", var_name))
    
    reduced <- lm(as.formula(base_formula), data = train_filtered)
    full    <- lm(full_formula, data = train_filtered)
    
    anova_result <- anova(reduced, full)
    
    tibble(
      Variable   = var_name,
      `RSS Drop` = round(anova_result$`Sum of Sq`[2], 4),
      `Adj R²`   = round(summary(full)$adj.r.squared, 4),
      `p-value`  = format.pval(anova_result$`Pr(>F)`[2], digits = 3),
      Levels     = length(unique(train_filtered[[var_name]]))
    )
  }) |> 
    arrange(desc(`RSS Drop`))
  
  print(results, n = Inf)
  invisible(results)
}

results <- evaluate_all_categorical()
```

#### Results

We'll use a 5 RSS drop threshold to determine which predictors to keep. This is somewhat arbitrary but provides a way to focus on predictors that are providing meaningful improvements to the model fit.

```{r}
Removed = union(
  Removed, c(
    'LandSlope', 'RoofStyle', 'YrSold', 'MoSold', 'RoofMatl', 
    'Fence', 'Condition2', 'LandContour', 'LotConfig', 'Alley', 
    'Heating', 'ExterCond', 'MasVnrType', 'LotShape', 'Condition1', 
    'PavedDrive', 'HouseStyle', 'Electrical', 'SaleType', 'Functional', 
    'SaleConditon', 'BldgType', 'FireplaceQu'
  )
)
```

### Multicollinearity Check

```{r}
train_filtered = train |> 
  slice(-c(524, 1299)) |> 
  select(-any_of(Removed))
```

Now down to 25 predictors. We will check for multicollinearity using VIF and remove any predictors with VIF > 5, starting with the most extreme values.

```{r}
model_mlr2 <- lm(log(SalePrice) ~ . - Exterior2nd, data = train_filtered) # Removing Exterior2nd due to high VIF and low RSS drop)
summary(model_mlr2)

par(mfrow = c(2, 2))
plot(model_mlr2)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(model_mlr2$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(model_mlr2$residuals), sd = sd(model_mlr2$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)

# show which variables are problematic
vif_values <- car::vif(model_mlr2)
vif_values |> 
  as_tibble(rownames = "Variable") |> 
  arrange(desc(`GVIF^(1/(2*Df))`)) |> 
  mutate(across(where(is.numeric), round, 3))
```

The model has some multicollinearity issues with GrLivArea, GarageArea, and TotalBsmtSF. However, these are some of the most important predictors in the model and removing them would significantly reduce the model's predictive power. We will consider combining these predictors into a single predictor.

Exterior2nd and TotRmsAbvGrd presented the next highest VIF values and were not providing meaningful improvements to the model fit, so we will remove them from the model.

```{r}
Removed = union(Removed, c('Exterior2nd', 'TotRmsAbvGrd'))

train_filtered = train |> 
  filter(!Id %in% c(524, 1299)) |> 
  select(-any_of(Removed)) |>
  mutate(TotalSF = GrLivArea + TotalBsmtSF + GarageArea) |> 
  select(-c(GrLivArea, TotalBsmtSF, GarageArea))

model_mlr2 <- lm(log(SalePrice) ~ ., data = train_filtered)
summary(model_mlr2)

par(mfrow = c(2, 2))
plot(model_mlr2)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(model_mlr2$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(model_mlr2$residuals), sd = sd(model_mlr2$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)

# show which variables are problematic
vif_values <- car::vif(model_mlr2)
vif_values |> 
  as_tibble(rownames = "Variable") |> 
  arrange(desc(`GVIF^(1/(2*Df))`)) |> 
  mutate(across(where(is.numeric), round, 3))

reduced = lm(log(SalePrice) ~ TotalSF, data = train_filtered)
anova(reduced, model_mlr2)
```

model_mlr2 still has significantly more predictive power than a single-variable TotalSF model. We'll proceed with stepwise starting with highest RSS drop variables and checking for multicollinearity at each step.

### Stepwise Feature Selection

We will forgo reviewing residual plots during this step until we begin transformations on predictor variables.

```{r}
model_mlr2.1 = lm(log(SalePrice) ~ TotalSF + Neighborhood, data = train_filtered)
summary(model_mlr2.1)
anova(reduced, model_mlr2.1)
vif(model_mlr2.1)
  
model_mlr2.2 = lm(log(SalePrice) ~ TotalSF + Neighborhood + OverallQual, data = train_filtered)
summary(model_mlr2.2)
anova(model_mlr2.1, model_mlr2.2)
vif(model_mlr2.2)

model_mlr2.3 = lm(log(SalePrice) ~ TotalSF + Neighborhood + OverallQual, data = train_filtered)
summary(model_mlr2.3)
anova(model_mlr2.2, model_mlr2.3)
vif(model_mlr2.3)

model_mlr2.4 = lm(log(SalePrice) ~ TotalSF + Neighborhood + OverallQual + MSSubClass, data = train_filtered)
summary(model_mlr2.4)
anova(model_mlr2.3, model_mlr2.4)
vif(model_mlr2.4)

model_mlr2.5 = lm(log(SalePrice) ~ TotalSF + Neighborhood + OverallQual + MSSubClass + KitchenQual, data = train_filtered)
summary(model_mlr2.5)
anova(model_mlr2.4, model_mlr2.5)
vif(model_mlr2.5)
```

Both MSSubClass and Kitchen Qual provided about 1 or 2 RSS drop and started to show some multicolinearity issues. This may be due to neighborhood having 25 levels and potentially capturing some of the same information. One option is to combine levels of neighborhood into fewer groups, but for now we will proceed with the model without these two predictors.


### Feature Transformations and Interactions

```{r}
model_mlr2.21 = lm(log(SalePrice) ~ TotalSF + Neighborhood + OverallQual + TotalSF:OverallQual, data = train_filtered)
anova(model_mlr2.2, model_mlr2.21)
vif(model_mlr2.21)
```

model_mlr2.21 starts to show signs of multicollinearity when adding Neighborhood or OverallQual interactions. Our model either needs to use difference predictors or we need to collapse from 25 levels in neighborhood to a few. We do not have enough information about the neighborhood to confidently collapse into groups, so we will proceed without interactions.

### Final Model and Assumptions Check

Previous review of residual plots showed showed some curviture in Q-Q Residuals and minor questions regarding normality. This is a final attempt to ensure assumptions of linear regression are met as closely as possible for our preferred model.

```{r}
model_mlr3 <- lm(log(SalePrice) ~ TotalSF + Neighborhood + OverallQual, data = train_filtered)
summary(model_mlr3)

par(mfrow = c(2, 2))
plot(model_mlr3)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(model_mlr3$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(model_mlr3$residuals), sd = sd(model_mlr3$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

We will start addressing normality problems by reviewing each predictor against SalePrice and transform to fit.

```{r}
library(gridExtra)

# Three plots to compare transformations
p1 <- ggplot(train_filtered, aes(x = TotalSF, y = log(SalePrice))) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "2. log(SalePrice) vs TotalSF",
       x = "TotalSF", y = "log(SalePrice)") +
  theme_bw(base_size = 12)

p2 <- ggplot(train_filtered, aes(x = log(TotalSF), y = log(SalePrice))) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "3. log(SalePrice) vs log(TotalSF)",
       x = "log(TotalSF)", y = "log(SalePrice)") +
  theme_bw(base_size = 12)

p3 <- ggplot(train_filtered, aes(x = log(TotalSF), y = SalePrice^2)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "3. log(SalePrice) vs TotalSF^2",
       x = "TotalSF^2", y = "log(SalePrice)") +
  theme_bw(base_size = 12)

p4 <- ggplot(train_filtered, aes(x = log(TotalSF), y = SalePrice^2)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "3. log(SalePrice) vs TotalSF^3/4",
       x = "TotalSF^3/4", y = "log(SalePrice)") +
  theme_bw(base_size = 12)

grid.arrange(p1, p2, p3, p4, ncol = 2)

# Three versions of TotalSF
model_linear <- lm(log(SalePrice) ~ TotalSF + Neighborhood + OverallQual, 
                   data = train_filtered)

model_log <- lm(log(SalePrice) ~ log(TotalSF) + Neighborhood + OverallQual, 
                data = train_filtered)

model_quad <- lm(log(SalePrice) ~ poly(TotalSF, 2) + Neighborhood + OverallQual, 
                 data = train_filtered)

par(mfrow = c(3, 3), mar = c(4, 4, 3, 1))

# Row 1: Linear TotalSF (current model_mlr3)
hist(model_linear$residuals, probability = TRUE, 
     main = "Linear TotalSF\nResidual Histogram", col = "lightblue")
curve(dnorm(x, mean = mean(model_linear$residuals), 
            sd = sd(model_linear$residuals)), 
      add = TRUE, col = "red", lwd = 2)
qqnorm(model_linear$residuals, main = "Linear: Q-Q Plot")
qqline(model_linear$residuals, col = "red")

# Row 2: log(TotalSF)
hist(model_log$residuals, probability = TRUE, 
     main = "log(TotalSF)\nResidual Histogram", col = "lightblue")
curve(dnorm(x, mean = mean(model_log$residuals), 
            sd = sd(model_log$residuals)), 
      add = TRUE, col = "red", lwd = 2)
qqnorm(model_log$residuals, main = "log: Q-Q Plot")
qqline(model_log$residuals, col = "red")

# Row 3: Quadratic TotalSF
hist(model_quad$residuals, probability = TRUE, 
     main = "Quadratic TotalSF\nResidual Histogram", col = "lightblue")
curve(dnorm(x, mean = mean(model_quad$residuals), 
            sd = sd(model_quad$residuals)), 
      add = TRUE, col = "red", lwd = 2)
qqnorm(model_quad$residuals, main = "Quadratic: Q-Q Plot")
qqline(model_quad$residuals, col = "red")

par(mfrow = c(1, 1))
```

A quadratic transformation may be slightly beneficial, but at the cost of interpretability. Since it does not fully address concerns of normality, we will proceed with caution. 

```{r}

# Boxplot of log(SalePrice) by Neighborhood (ordered by median)
ggplot(train_filtered, aes(x = reorder(Neighborhood, log(SalePrice), FUN = median), 
                           y = log(SalePrice))) +
  geom_boxplot(outlier.alpha = 0.6, fill = "steelblue", alpha = 0.7) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3, size = 1) +
  labs(title = "log(SalePrice) by Neighborhood (ordered by median)",
       subtitle = "Clear differences between neighborhoods",
       x = "Neighborhood",
       y = "log(SalePrice)") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9))

par(mfrow = c(2, 2), mar = c(4.5, 4, 3, 1))

# Residual histogram
hist(model_linear$residuals, probability = TRUE,
     main = "Overall Residual Histogram", col = "lightblue")
curve(dnorm(x, mean = mean(model_linear$residuals),
            sd = sd(model_linear$residuals)),
      add = TRUE, col = "red", lwd = 2)

# Q-Q plot
qqnorm(model_linear$residuals, main = "Q-Q Plot of Residuals")
qqline(model_linear$residuals, col = "red", lwd = 2)

# Boxplot of residuals by Neighborhood
boxplot(model_linear$residuals ~ train_filtered$Neighborhood,
        main = "Residuals by Neighborhood",
        xlab = "Neighborhood", ylab = "Residuals",
        col = "lightblue")
abline(h = 0, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

Unlike TotalSF, Neighborhood is categorical and will not benefit from mathmatical transformations. There are very large differences between neighborhoods. Some neighborhoods have much higher median SalePrice than others, and the variance of SalePrice also differs between neighborhoods. This may be contributing to the non-normality of residuals and heteroscedasticity. Grouping levels may effect our predictive power, so our final model will include Neighborhood as-is and a reduced model with Neighborhood grouped into fewer levels for comparison.

```{r}
# Create a grouped version of Neighborhood (4 logical tiers)
train_filtered <- train_filtered |>
  mutate(
    NeighborhoodGroup = case_when(
      Neighborhood %in% c("NoRidge", "NridgHt", "StoneBr", "Veenker", "Timber") ~ "HighEnd",
      Neighborhood %in% c("CollgCr", "Crawfor", "ClearCr", "Somerst", "Gilbert", "NWAmes", "SawyerW", "Blmngtn") ~ "UpperMid",
      Neighborhood %in% c("NAmes", "Mitchel", "NPkVill", "SWISU", "Sawyer", "Edwards") ~ "Middle",
      Neighborhood %in% c("BrkSide", "OldTown", "IDOTRR", "MeadowV", "BrDale", "Blueste")~ "Lower",
      TRUE ~ "Other"
    )
  )

# Convert to ordered factor
train_filtered$NeighborhoodGroup <- factor(
  train_filtered$NeighborhoodGroup, levels = c("Lower", "Middle", "UpperMid", "HighEnd")
)

# Fit the new model with grouped neighborhoods
model_mlr3_normal <- lm(log(SalePrice) ~ TotalSF + NeighborhoodGroup + OverallQual, data = train_filtered)

# Compare the two models
summary(model_mlr3_normal)
cat("\n=== MODEL COMPARISON ===\n")
cat("model_mlr3 (25 levels):     Adj R² =", round(summary(model_mlr3)$adj.r.squared, 4),
    " | AIC =", round(AIC(model_mlr3), 1), "\n")
cat("model_mlr3_normal (4 groups): Adj R² =", round(summary(model_mlr3_normal)$adj.r.squared, 4),
    " | AIC =", round(AIC(model_mlr3_normal), 1), "\n")

anova(model_mlr3, model_mlr3_normal)

par(mfrow = c(2, 2))
plot(model_mlr3_normal)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(
  model_mlr3_normal$residuals, 
  probability = TRUE, # Use density scale
  main = "Histogram of Residuals",
  xlab = "Residuals",
  col = "lightblue",
  border = "black"
)

# Add normal curve
curve(
  dnorm(
    x, 
    mean = mean(model_mlr3_normal$residuals), 
    sd = sd(model_mlr3_normal$residuals)
  ), 
    add = TRUE, 
    col = "red", 
    lwd = 2
)
```

There's evidence that grouping Neighborhood into 4 groups results in a significant drop in model fit while not significantly improving normality judging from visual residual plots (partial f-test p-value <.0001). The model with 25 levels of Neighborhood has better fit and predictive power.

Finally we'll address any remaining outliers' leverage and Cook's distance then refit the model if needed. 

```{r}
# Focus on your best model (model_mlr3)
p <- length(coef(model_mlr3))           # number of parameters
n <- nrow(train_filtered)
leverage_threshold <- 2 * p / n         # cutoff

cat("Leverage threshold (2p/n):", round(leverage_threshold, 4), "\n\n")

# Extract leverage and Cook's distance
hat_values <- hatvalues(model_mlr3)
cooks_d <- cooks.distance(model_mlr3)

# Find high leverage points
high_leverage_idx <- which(hat_values > leverage_threshold)

cat("Found", length(high_leverage_idx), "high leverage points\n\n")

# Show details of those observations
if(length(high_leverage_idx) > 0) {
  high_lev_points <- train_filtered[high_leverage_idx, ] |>
    select(TotalSF, Neighborhood, OverallQual, SalePrice) |>
    mutate(
      Row = high_leverage_idx,
      Leverage = round(hat_values[high_leverage_idx], 4),
      Cooks_D = round(cooks_d[high_leverage_idx], 5),
      Residual = round(model_mlr3$residuals[high_leverage_idx], 4)
    ) |>
    arrange(desc(Leverage))
  
  print(high_lev_points)
}
```

Our data does not have outliers with > .5 Cook's distance, but it does have some high leverage points.

The top 4 leverage outliers are still meaningful, representing abnormal sales conditions and unique properties. 

All NPkVill obersvations show as high leverage points, but this is a unique neighborhood. This is likely a real phenomenon and not an error, so we will keep these observations in the model.

## Model Comparison and Evaluation

SLR = `(SalePrice | GrLivArea)`
MLR1 = `(SalePrice | GrLivArea + FullBath)`
MLR2 = `(SalePrice | 25 predictors)` full model after EDA considering linearily related variables
MLR3 = `(SalePrice | TotalSF + Neighborhood + OverallQual)`
MLR3.N = `(SalePrice | TotalSF + NeighborhoodGroup + OverallQual)` After grouping Neighborhood into 4 groups based on median sales price.

- **Adjusted R²**: How well the model explains variance (higher is better)
- **PRESS**: Leave-one-out cross-validation error (lower is better)
- **CV RMSE (log scale)**: 10-fold cross-validation root mean squared error (lower is better)
- **AIC**: Balance between fit and complexity (lower is better)

```{r, warning=FALSE, message=FALSE}
library(caret)
library(knitr)

# Define cross-validation method
cv_control <- trainControl(method = "cv", number = 10, savePredictions = TRUE)

# Function to get 10-fold CV RMSE on log(SalePrice) scale
cv_rmse <- function(model, data) {
  train(log(SalePrice) ~ ., 
        data = data, 
        method = "lm", 
        trControl = cv_control)$results$RMSE
}

# PRESS function (Leave-One-Out Cross Validation statistic)
press_stat <- function(model) {
  res <- residuals(model)
  h   <- hatvalues(model)
  sum((res / (1 - h))^2, na.rm = TRUE)
}

# Model Comparison Table
model_comparison <- tibble(
  Model                  = c("SLR", 
                            "MLR1", 
                            "MLR2", 
                            "MLR3",
                            "MLR3.N"),
  
  `Adjusted R²`          = c(
    round(summary(model_slr)$adj.r.squared, 4),
    round(summary(model_mlr1)$adj.r.squared, 4),
    round(summary(model_mlr2)$adj.r.squared, 4),
    round(summary(model_mlr3)$adj.r.squared, 4),
    round(summary(model_mlr3_normal)$adj.r.squared, 4)
  ),
  
  PRESS                  = c(
    round(press_stat(model_slr), 0),
    round(press_stat(model_mlr1), 0),
    round(press_stat(model_mlr2), 0),
    round(press_stat(model_mlr3), 0),
    round(press_stat(model_mlr3_normal), 0)
  ),
  
  `CV RMSE (log)`  = c(
    round(cv_rmse(model_slr, train_filtered), 4),
    round(cv_rmse(model_mlr1, train_filtered), 4),
    round(cv_rmse(model_mlr2, train_filtered), 4),
    round(cv_rmse(model_mlr3, train_filtered), 4),
    round(cv_rmse(model_mlr3_normal, train_filtered), 4)
  ),
  
  AIC                    = c(
    round(AIC(model_slr), 1),
    round(AIC(model_mlr1), 1),
    round(AIC(model_mlr2), 1),
    round(AIC(model_mlr3), 1),
    round(AIC(model_mlr3_normal), 1)
  ),
  
  `Kaggle Score`      = c("---", "---", "---", "---", "---")

)

model_comparison |> 
  kable(format = "markdown", 
        caption = "Model Comparison Table (PRESS + 10-fold CV)",
        align = "lrrrr")
```

Model MLR3 has the best performance across all metrics, with the highest adjusted R², lowest PRESS, lowest CV RMSE, and lowest AIC. The model with grouped Neighborhood (MLR3.N) has slightly worse performance than MLR3 but better AIC, suggesting that the detailed neighborhood information is important enough to justify the added complexity.

## Kaggle Submission

First need to mutate the test dataset to have the same predictors as our final model. We will also need to impute any missing values in the test set using the same methods as the training set.

```{r}
test = test |> 
  mutate(TotalSF = GrLivArea + TotalBsmtSF + GarageArea)

test <- test |>
  mutate(
    NeighborhoodGroup = case_when(
      Neighborhood %in% c("NoRidge", "NridgHt", "StoneBr", "Veenker", "Timber") ~ "HighEnd",
      Neighborhood %in% c("CollgCr", "Crawfor", "ClearCr", "Somerst", "Gilbert", "NWAmes", "SawyerW", "Blmngtn") ~ "UpperMid",
      Neighborhood %in% c("NAmes", "Mitchel", "NPkVill", "SWISU", "Sawyer", "Edwards") ~ "Middle",
      Neighborhood %in% c("BrkSide", "OldTown", "IDOTRR", "MeadowV", "BrDale", "Blueste")~ "Lower",
      TRUE ~ "Other"
    )
  )

# Convert to ordered factor
test$NeighborhoodGroup <- factor(
  test$NeighborhoodGroup, levels = c("Lower", "Middle", "UpperMid", "HighEnd")
)
```


```{r, eval=FALSE}
# Function to create Kaggle submission
create_kaggle_submission <- function(model, filename = "submission.csv") {
  
  # Align factor levels to prevent "new levels" error
  test_processed <- test
  for (var in names(model$xlevels)) {
    if (var %in% names(test_processed)) {
      test_processed[[var]] <- factor(test_processed[[var]], 
                                      levels = model$xlevels[[var]])
    }
  }
  
  # Generate predictions on log scale
  test_pred <- predict(model, newdata = test_processed, interval = "none")
  
  # Back-transform to original scale
  saleprice <- exp(test_pred)
  
  # Count and replace any NAs with median
  na_count <- sum(is.na(saleprice))
  if (na_count > 0) {
    median_price <- median(saleprice, na.rm = TRUE)
    saleprice[is.na(saleprice)] <- median_price
    cat("→ Replaced", na_count, "NA predictions with median SalePrice =", 
        round(median_price, 2), "\n")
  }
  
  # Create submission in exact format as sample_submission.csv
  submission <- tibble(
    Id        = test$Id,
    SalePrice = as.double(saleprice)   # ensure dbl type
  )
  
  write_csv(submission, here("Data", filename))
  
  cat("Submission saved to: Data/", filename, "\n")
  cat("Rows:", nrow(submission), 
      " | Final NAs in SalePrice:", sum(is.na(submission$SalePrice)), "\n\n")
  
  invisible(na_count)
}

create_kaggle_submission(model_slr, "submission_slr.csv")
create_kaggle_submission(model_mlr1, "submission_mlr1.csv")
create_kaggle_submission(model_mlr2, "submission_mlr2.csv")
create_kaggle_submission(model_mlr3, "submission_mlr3.csv")
create_kaggle_submission(model_mlr3_normal, "submission_mlr3n.csv")
```

Submission saved to: Data/ submission_slr.csv 
Rows: 1459  | Final NAs in SalePrice: 0 

Submission saved to: Data/ submission_mlr1.csv 
Rows: 1459  | Final NAs in SalePrice: 0 

→ Replaced 9 NA predictions with median SalePrice = 155986.3 
Submission saved to: Data/ submission_mlr2.csv 
Rows: 1459  | Final NAs in SalePrice: 0 

→ Replaced 2 NA predictions with median SalePrice = 157249.2 
Submission saved to: Data/ submission_mlr3.csv 
Rows: 1459  | Final NAs in SalePrice: 0 

→ Replaced 2 NA predictions with median SalePrice = 157175.4 
Submission saved to: Data/ submission_mlr3n.csv 
Rows: 1459  | Final NAs in SalePrice: 0 