---
title: "Analysis 2: Predictive Modeling for House Prices"
description: "This analysis focuses on building predictive models to forecast future trends based on historical data.
author: "Aaron Powell & Joshua Samuel"
date: r.Sys.Date()
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(GGally)
library(car)
source("code/utils_data_prep.R") # Data cleaned and prepped for LR
```

## Simple Linear Regression

Taking a simple approach first, we will model log(SalePrice) as a function of Above grade (ground) living area square feet (GrLivArea). GrLivArea is a strong predictor of SalePrice and is commonly used in housing price models.

```{r}
# Compute correlation matrix from numeric variables only
numeric_train <- train |> select(where(is.numeric))
cor_matrix <- cor(numeric_train, use = "complete.obs")

saleprice_cor <- cor_matrix[, "SalePrice"]
cor_df <- tibble(
  Variable = names(saleprice_cor),
  Correlation = round(saleprice_cor, 3),
  Abs_Correlation = abs(Correlation)
) |>
  arrange(desc(Abs_Correlation))
print("Top variables by |correlation| with SalePrice:")
print(cor_df, n = 10)
```

```{r}
model_slr <- lm(log(SalePrice) ~ GrLivArea, data = train, subset = c(-1299, -524)) # Exclude outlier 1299 and 524
summary(model_slr)
par(mfrow = c(2, 2))
plot(model_slr)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(model_slr$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(model_slr$residuals), sd = sd(model_slr$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

1299 remains an outlier. It is a 5600 sqft house sold as a unfinished build for 160,000. This is an extreme case and likely not representative of the general market, so we will exclude it from our modeling.

524 is a 4700 sqft house sold for 184750 as a partial unfinished new build, which represents another extreme case. For SLR we are not considering any other factor than GrLivArea so we will exclude this special case.

## Initial Multiple Linear Regression

Number of Full Bathrooms (FullBath) is another highly correlated predictor of SalePrice. We will add it to the model to see if it improves fit and addresses any issues with the SLR model.

```{r}
model_mlr1 <- lm(log(SalePrice) ~ GrLivArea + FullBath, data = train, subset = c(-1299, -524, -636, -1124))
summary(model_mlr1)

par(mfrow = c(2, 2))
plot(model_mlr1)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(model_mlr1$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(model_mlr1$residuals), sd = sd(model_mlr1$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

636 is a 3395 sqft house sold for 200,000 as a abnormal sale. This is a large house sold for a very low price. Since abnormal sales can be anything from forclosure to short sale or trade, we may consider excluding this group. For this model we will only exclude the one outlier. 

Only explaining another 3% of the variance, this model is not a significant improvement over the SLR. The residual plots show similar patterns to the SLR, with some improvement in homoscedasticity but still some curvature in the residuals vs fitted plot. The histogram of residuals looks more normal. 

## Enhanced Multiple Linear Regression Models

Starting with a full model including all predictors and then we will work down to a more parsimonious model. 

```{r}
full_model <- lm(log(SalePrice) ~ ., data = train)
summary(full_model)
par(mfrow = c(2, 2))
plot(full_model)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(full_model$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(full_model$residuals), sd = sd(full_model$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

Histogram of residuals looks very condensed to the center and long tailed. Clearly not normal. Studentized residuals show some curvature in the residuals vs fitted plot, but overall the model seems to be doing a good job of capturing the variance in SalePrice with the log transformation.

The model displays overfit and normality concerns. We will work on improving the model by removing some predictors and adding interactions and transformations.

### Linearity Check

First we will look for linearity among numerical predictors.

```{r}
# Get numeric predictors (exclude SalePrice)

# Exclude the two extreme outliers identified earlier
train_filtered <- train |> 
  slice(-c(524, 1299)) # These two really skew the relationships and are not representative of the general market

numeric_predictors <- train_filtered |> 
  select(where(is.numeric), -SalePrice) |> 
  names()

# Reshape to long format for faceting
train_long <- train_filtered |> 
  pivot_longer(
    cols = all_of(numeric_predictors),
    names_to = "Predictor",
    values_to = "Value"
  )

# Create the linearity diagnostic plot
ggplot(train_long, aes(x = Value, y = log(SalePrice))) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", color = "red", linewidth = 1, se = FALSE) +
  facet_wrap(~ Predictor, scales = "free_x", ncol = 4) +
  labs(
    title = "Linearity Check: Numeric Predictors vs log(SalePrice)",
    subtitle = "Red line = loess smoother. Look for straight lines (good linearity).",
    x = NULL,
    y = "log(SalePrice)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 8)
  )
```

Reviewing the scatterplots of numeric data against log(SalePrice) with loess smoothers:

GrLiveArea shows a possible quadratic relationship, with diminishing returns at a higher value. 

X1stFlrSF and X2ndFlrSf show linear relationships. But this may already be captured by GrLivArea, which is the sum of 1st and 2nd floor area. We may consider removing these two predictors to address multicollinearity.

```{r}
reduced <- lm(log(SalePrice) ~ GrLivArea, data = train_filtered)
full <- lm(log(SalePrice) ~ GrLivArea + X1stFlrSF + X2ndFlrSF, data = train_filtered)
anova(reduced, full)
vif(full)
```

While there is evidence to suggest 1stFlrSF and 2ndFlrSF improve the model in a statistical sense, multicollinearity occurs because GrLIvArea is effectively of sum of the two variables. 

TotRmsAbvGrd show possible linearity but could be capturing the same information as GrLivArea, so we may consider removing it to address multicollinearity.

```{r}
full <- lm(log(SalePrice) ~ GrLivArea + TotRmsAbvGrd, data = train_filtered)
anova(reduced, full)
vif(full)
```

There is overwhelming evidence adding TotRmsAbvGrd to a model containing GrLivArea produces meaningful predictive power(parital f-test p-value <.0001). Therefore we will keep TotRmsAbvGrd in the model. 

LotArea shows a possible linear relationship, with diminishing returns at lower values and an outlier at the high end or possible quadratic relationship at extremes. 

```{r}
full <- lm(log(SalePrice) ~ GrLivArea + LotArea, data = train_filtered)
anova(reduced, full)
vif(full)
```

There is overwhelming evidence adding LotArea produces meaningful predictive power(parital f-test p-value <.0001). Therefore we will keep LotArea in the model.

BSmtFinSf1 and TotalBsmtSF show possible linear relationship with one outlier at the high end. We may consider grouping all basement variables and testing for lack of fit against a model with TotalBsmtSF as the only basement predictor.

```{r}
full = lm(log(SalePrice) ~ GrLivArea + TotalBsmtSF, data = train_filtered)
anova(reduced, full)
vif(full)
```

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea + TotalBsmtSF, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + TotalBsmtSF + BsmtQual + BsmtExposure, data = train_filtered)
anova(reduced, full)
vif(full)
```

There's overwhelming evidence to suggest TotalBsmtSF in a model with GrLivArea has meaningful predictive power (partial f-test p-value <.0001). Any attempts to add other bsmt variables results in high multicolinearity. 

```{r}
feature_selection = c(feature_selection, 'TotalBsmtSF')
```

GarageArea shows a possible linear relationship with log(SalePrice) but with some outliers at the high end. We may consider grouping all garage variables and testing for lack of fit against a model with GarageArea as the only garage predictor.

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + GarageArea, data = train_filtered)
anova(reduced, full)
vif(full)
```

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea + GarageArea, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + GarageArea + GarageType, data = train_filtered)
anova(reduced, full)
vif(full)
```

There's overwhelming evidence to suggest GarageArea in a model with GrLivArea and GarageType has meaningful predictive power (partial f-test p-value <.0001). Any attempts to add other Garage variables other than GarageType results in high multicolinearity. 

FirePlaces show possible linear relationship with log(SalePrice) but may not be enough to provide predictive power. We may consider grouping all fireplace variables and testing for lack of fit against a model with FirePlaces as the only fireplace predictor.

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + Fireplaces, data = train_filtered)
anova(reduced, full)
vif(full)
```

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea + Fireplaces, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + Fireplaces + FireplaceQu, data = train_filtered)
anova(reduced, full)
vif(full)
```

There's overwhelming evidence to suggest Fireplaces in a model with GrLivArea and FireplaceQu has meaningful predictive power (partial f-test p-value <.0001). 

While WoodDeckSF has potential linear relationship, the variance is too high while linearity is nearly flat. We may consider removing this predictor.

We saw earliier in model_mlr1 that FullBath has a linear relationship with log(SalePrice) and improved our model by adding 3% to the R². We will keep it in the model unless it shows multicollinearity with other predictors.

```{r}
reduced = lm(log(SalePrice) ~ GrLivArea + FullBath, data = train_filtered)
full = lm(log(SalePrice) ~ GrLivArea + FullBath + HalfBath, data = train_filtered)
anova(reduced, full)
vif(full)
```

There's evidence to suggest Halfbath may have some predictive power (partial f-test p-value = 0.0002).

Flat relationships with log(SalePrice) are observed and will be removed from our model:
BedroomAbvGr, BsmtFullBath, BsmtHalfBath, EnclosedPorch, HalfBath, KitchenAbvGr, LowQualFinSf, MiscVal, OpenPorchSF, PoolArea, ScreenPorch, X3SsnPorch, YearBuilt.

Considering these values have potential linearity, but too strong of grouping at 0, we may consider using binary predictors: MasVnrArea, WoodDeckSf

LotFrontage has imputed values that may be skewing the relationship. We will proceed with caution. 

```{r}
Removed = union(Removed, c('X1stFlrSF', 'X2ndFlrSF', 'BsmtQual', 'BsmtCond', 'BsmtExposure', 'BsmtFinType1', 'BsmtFinSF1', 'BsmtFinType2', 'BsmtFinSF2', 'BsmtUnfSF', 'BsmtFullBath', 'BsmtHalfBath', 'GarageFinish', 'GarageQual', 'GarageCond', 'WoodDeckSF', 'BedroomAbvGr', 'BsmtFullBath', 'BsmtHalfBath', 'EnclosedPorch', 'KitchenAbvGr', 'LowQualFinSf', 'MiscVal', 'OpenPorchSF', 'PoolArea', 'ScreenPorch', 'X3SsnPorch', 'YearBuilt'))

train_filtered = train |> 
  slice(-c(524, 1299)) |> 
  select(-any_of(Removed))
```

left off at this point. Ready to look at box plots for categorical predictors and visually inspect for linearity and potential interactions.



```{r}
vif_values <- vif(model_mlr2)
print(vif_values[vif_values > 5]) # start with the most extreme VIF values and work down
```


outline




# 6. Model 3+: Enhanced Multiple Linear Regression Models -------------
# - Build at least one (ideally 2–3) improved models using techniques from class:
#     - Add meaningful predictors (e.g., OverallQual, HouseAge, TotalSF, Neighborhood as factor/dummies, ExterQual, etc.)
#     - Include transformations (log(GrLivArea), sqrt(LotArea), etc.)
#     - Add interactions (e.g., GrLivArea * OverallQual, GrLivArea * Neighborhood, YearBuilt * Remod)
#     - Feature selection: forward/backward/stepwise (stepAIC or manual), or correlation/VIF screening
# - Example enhanced model: log(SalePrice) ~ log(GrLivArea) + OverallQual + HouseAge + TotalBath + Neighborhood + GrLivArea:OverallQual + ...
# - Fit each as model_mlr2, model_mlr3, etc.
# - Comment heavily on why each variable/interaction was added.

# 7. Model Diagnostics and Assumption Checking (Critical!) ------------
# For EVERY model (SLR, MLR1, and enhanced):
#   a. Residual plots:
#      - Residuals vs. Fitted (check linearity & homoscedasticity)
#      - Normal Q-Q plot (normality of residuals)
#      - Scale-Location plot
#      - Residuals vs. Leverage (outliers/influential points)
#   b. Influential point analysis:
#      - Cook's Distance (> 4/n threshold)
#      - Leverage (hat values > 2p/n)
#      - DFFITS, DFBETAS if needed
#   c. Multicollinearity: VIF (variance inflation factors) – address if > 5–10
#   d. Address violations:
#      - Transformations already in model?
#      - Remove influential points? (justify with sensitivity analysis)
#      - Add polynomial terms or interactions if curvature present?
# - Create diagnostic plot functions or loops to avoid repetition.

# 8. Model Comparison and Evaluation ----------------------------------
# - Create a comparison table (as required in the project description):
#   Columns: Model, Adjusted R², CV PRESS (use PRESS statistic or cvTools/ caret for cross-validation), AIC, Kaggle Score (RMSE on log scale)
# - Compute internal CV PRESS (e.g., via leave-one-out or k-fold)
# - Generate predictions on test set for Kaggle:
#     - Predict log(SalePrice), exponentiate back to SalePrice
#     - Create submission.csv with Id and SalePrice
#     - Note the public Kaggle score for each model
# - Clearly state which model is "best" and why (balance of fit, assumptions, predictive performance, parsimony).

# 9. Best Model Selection and Kaggle Submission -----------------------
# - Export the final submission file(s)
# - Optional: Save model objects (RDS) for Shiny app or report reproducibility

# 10. Session Info and Reproducibility --------------------------------
# sessionInfo()
# Save any plots/tables for the report appendix
