---
title: "Analysis 2: Predictive Modeling for House Prices"
description: "This analysis focuses on building predictive models to forecast future trends based on historical data.
author: "Aaron Powell & Joshua Samuel"
date: r.Sys.Date()
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(GGally)
library(car)
source("code/utils_data_prep.R") # Data cleaned and prepped for LR
```

## Simple Linear Regression

Taking a simple approach first, we will model log(SalePrice) as a function of Above grade (ground) living area square feet (GrLivArea). GrLivArea is a strong predictor of SalePrice and is commonly used in housing price models.

```{r}
saleprice_cor <- cor_matrix[, "SalePrice"]
cor_df <- tibble(
  Variable = names(saleprice_cor),
  Correlation = round(saleprice_cor, 3),
  Abs_Correlation = abs(Correlation)
) |>
  arrange(desc(Abs_Correlation))
print("Top variables by |correlation| with SalePrice:")
print(cor_df, n = 10)
```

```{r}
model_slr <- lm(log(SalePrice) ~ GrLivArea, data = train, subset = c(-1299, -524)) # Exclude outlier 1299 and 524
summary(model_slr)
par(mfrow = c(2, 2))
plot(model_slr)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(model_slr$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(model_slr$residuals), sd = sd(model_slr$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

1299 remains an outlier. It is a 5600 sqft house sold as a unfinished build for 160,000. This is an extreme case and likely not representative of the general market, so we will exclude it from our modeling.

524 is a 4700 sqft house sold for 184750 as a partial unfinished new build, which represents another extreme case. For SLR we are not considering any other factor than GrLivArea so we will exclude this special case.

## Initial Multiple Linear Regression

## Full Multiple Linear Regression

We will proceed with log(SalePrice) since Analysis 1 displayed some heteroscedasticity. 

```{r}
full_model <- lm(log(SalePrice) ~ ., data = train)
par(mfrow = c(2, 2))
plot(full_model)
par(mfrow = c(1, 1))
```




The summary of the full model NA values for several predictors. This is likely due to perfect multicollinearity (e.g., YearBuilt and YearRemodAdd are highly correlated). 

```{r}
vif_values <- vif(full_model)
print(vif_values[vif_values > 5]) # start with the most extreme VIF values and work down
```


outline

# - Create key derived features (examples relevant to our course):
#     - log_SalePrice (for modeling and Kaggle submission)
#     - HouseAge = YrSold - YearBuilt
#     - TotalSF = GrLivArea + TotalBsmtSF
#     - TotalBath = FullBath + HalfBath + BsmtFullBath*0.5 + BsmtHalfBath*0.5
#     - Quality interactions or dummy variables for high-impact categoricals (e.g., OverallQual, Neighborhood, ExterQual, KitchenQual)
#     - Any other feature creation justified by domain knowledge or EDA (e.g., log transformations on skewed continuous vars like LotArea, GrLivArea).
# - Filter or subset if needed, but remember: Analysis 2 uses ALL neighborhoods.

# 3. Exploratory Data Analysis (EDA) for Modeling --------------------
# - Summary statistics and correlations (focus on SalePrice and key predictors).
# - Visualizations: scatterplots (SalePrice vs. GrLivArea, colored by key factors), histograms of SalePrice (and log(SalePrice)), boxplots by categorical vars.
# - Identify potential transformations and interactions early.

# 4. Model 1: Simple Linear Regression (Baseline) ---------------------
# - Formula: log(SalePrice) ~ GrLivArea   (or SalePrice if you prefer; discuss choice)
# - Fit with lm()
# - Store model object (e.g., model_slr)
# - Basic summary and initial diagnostics

# 5. Model 2: Required Multiple Linear Regression (MLR1) --------------
# - Formula: log(SalePrice) ~ GrLivArea + FullBath
# - Fit with lm()
# - Store as model_mlr1

# 6. Model 3+: Enhanced Multiple Linear Regression Models -------------
# - Build at least one (ideally 2–3) improved models using techniques from class:
#     - Add meaningful predictors (e.g., OverallQual, HouseAge, TotalSF, Neighborhood as factor/dummies, ExterQual, etc.)
#     - Include transformations (log(GrLivArea), sqrt(LotArea), etc.)
#     - Add interactions (e.g., GrLivArea * OverallQual, GrLivArea * Neighborhood, YearBuilt * Remod)
#     - Feature selection: forward/backward/stepwise (stepAIC or manual), or correlation/VIF screening
# - Example enhanced model: log(SalePrice) ~ log(GrLivArea) + OverallQual + HouseAge + TotalBath + Neighborhood + GrLivArea:OverallQual + ...
# - Fit each as model_mlr2, model_mlr3, etc.
# - Comment heavily on why each variable/interaction was added.

# 7. Model Diagnostics and Assumption Checking (Critical!) ------------
# For EVERY model (SLR, MLR1, and enhanced):
#   a. Residual plots:
#      - Residuals vs. Fitted (check linearity & homoscedasticity)
#      - Normal Q-Q plot (normality of residuals)
#      - Scale-Location plot
#      - Residuals vs. Leverage (outliers/influential points)
#   b. Influential point analysis:
#      - Cook's Distance (> 4/n threshold)
#      - Leverage (hat values > 2p/n)
#      - DFFITS, DFBETAS if needed
#   c. Multicollinearity: VIF (variance inflation factors) – address if > 5–10
#   d. Address violations:
#      - Transformations already in model?
#      - Remove influential points? (justify with sensitivity analysis)
#      - Add polynomial terms or interactions if curvature present?
# - Create diagnostic plot functions or loops to avoid repetition.

# 8. Model Comparison and Evaluation ----------------------------------
# - Create a comparison table (as required in the project description):
#   Columns: Model, Adjusted R², CV PRESS (use PRESS statistic or cvTools/ caret for cross-validation), AIC, Kaggle Score (RMSE on log scale)
# - Compute internal CV PRESS (e.g., via leave-one-out or k-fold)
# - Generate predictions on test set for Kaggle:
#     - Predict log(SalePrice), exponentiate back to SalePrice
#     - Create submission.csv with Id and SalePrice
#     - Note the public Kaggle score for each model
# - Clearly state which model is "best" and why (balance of fit, assumptions, predictive performance, parsimony).

# 9. Best Model Selection and Kaggle Submission -----------------------
# - Export the final submission file(s)
# - Optional: Save model objects (RDS) for Shiny app or report reproducibility

# 10. Session Info and Reproducibility --------------------------------
# sessionInfo()
# Save any plots/tables for the report appendix
